---
title: "Introduction to {rtables.officer}"
author: "Davide Garolini and Joe Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to {rtables.officer}}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---


```{r, include = FALSE}
suggested_dependent_pkgs <- c("dplyr")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = all(vapply(
    suggested_dependent_pkgs,
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
)
```
```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

## Introduction

The `rtables` package provides a framework to create, tabulate, and
output tables in R. Most of the design requirements for `rtables`
have their origin in studying tables that are commonly used to report
analyses from clinical trials; however, we were careful to keep
`rtables` a general purpose toolkit.

In this vignette, we give a short introduction into creating tables with `rtables` and exporting them into `docx` files with `rtables.officer`.

The content of this vignette is based on the following two resources:

* The [`rtables` useR 2020 presentation](https://www.youtube.com/watch?v=CBQzZ8ZhXLA)
by Gabriel Becker
* [`rtables` - A Framework For Creating Complex Structured Reporting Tables Via
Multi-Level Faceted Computations](https://arxiv.org/pdf/2306.16610).

The packages used in this vignette are `rtables.officer`, `rtables`, and `dplyr`:

```{r, message=FALSE}
library(rtables.officer)
library(dplyr)
```

## Overview

The `rtables` R package was designed to create and display complex
tables with R. The cells in an `rtable` may contain any high-dimensional
data structure which can then be displayed with cell-specific formatting
instructions. Currently, `rtables` can output tables in `ascii`, `html`, and 
`pdf` formats.

The `rtables.officer` package is designed to support export formats related to 
the Microsoft Office software suite, including Microsoft Word (`docx`) and 
Microsoft PowerPoint (`pptx`).

For a detailed guide on getting started with `rtables`, please refer to [this page](https://insightsengineering.github.io/rtables/latest-release/articles/rtables.html).

## Data

The data used in this vignette is a made up using random number
generators. The data content is relatively simple: one row per
imaginary person and one column per measurement: study arm, the
country of origin, gender, handedness, age, and weight.

```{r data}
n <- 400

set.seed(1)

df <- tibble(
  arm = factor(sample(c("Arm A", "Arm B"), n, replace = TRUE), levels = c("Arm A", "Arm B")),
  country = factor(sample(c("CAN", "USA"), n, replace = TRUE, prob = c(.55, .45)), levels = c("CAN", "USA")),
  gender = factor(sample(c("Female", "Male"), n, replace = TRUE), levels = c("Female", "Male")),
  handed = factor(sample(c("Left", "Right"), n, prob = c(.6, .4), replace = TRUE), levels = c("Left", "Right")),
  age = rchisq(n, 30) + 10
) %>% mutate(
  weight = 35 * rnorm(n, sd = .5) + ifelse(gender == "Female", 140, 180)
)

head(df)
```

Note that we use factor variables so that the level order is
represented in the row or column order when we tabulate the
information of `df` below.

## Building a Table

The aim of this vignette is to build the following table and then use `rtables.officer` to export it to a `docx` file.

```{r tbl, echo=FALSE}
lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by("arm") %>%
  split_cols_by("gender") %>%
  split_rows_by("country") %>%
  summarize_row_groups() %>%
  split_rows_by("handed") %>%
  summarize_row_groups() %>%
  analyze("age", afun = mean, format = "xx.xx")

tbl <- build_table(lyt, df)
tbl
```

First, we build the table as follows:

```{r tbl, echo=TRUE}
```

## Exporting to a `docx` File

Next, we will export the `rtables` object generated above (`tbl`) into a `docx` file, namely `example.docx`.

```{r, eval=FALSE}
output_dir <- "." # specify here where the file should be saved
export_as_docx(tbl, file = file.path(output_dir, "example.docx"))
```

The `export_as_docx()` function used above first converts the `rtables` object into a `flextable` object and then uses the `officer` package to save the output to a `docx` file.
